Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-09 paper/paper.tex
Add related work about gluing

\paragraph{Axioms}
Our Coq development assumes two axioms: functional extensionality
and propositional extensionality. The former is also required by
the Autosubst 2 libraries. Both axioms are known to be consistent
with Coq's metatheory.
%
These axioms bridge the gap between our mechanization and our informal
proofs. For example, in set theory, to show that two sets $[[S0]]$ and
$[[S1]]$ are equal, it suffices to show the extensional property that
$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
%
However, in Coq, sets of terms ($[[PowerSet STm]]$) are encoded as the 
type \texttt{tm
  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
not come with the extensionality property. Given two predicates $P$ and $Q$,
we cannot conclude that $P = Q$ when given a proof of
$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
predicate extensionality, an immediate corollary from functional
extensionality and propositional extensionality.

\paragraph{Encoding the logical relation in Coq}
We next discuss specific details of the Coq encoding of the logical relation
presented in Section~\ref{sec:logreldep}.

@@

\paragraph{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, supported by the tools
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

@@

More recent work such as \citet{Abel12}
and \citet{abel2008betaeta} make their definitions more explicit and
precise and thus more directly encodable in proof assistants.
Our logical relation resembles their definition of a
semantic universe hierarchy, although we close our relation
under expansion with respect to parallel reduction rather
than weak-head reduction. \scw{Why is this important?} \yl{if you are
  referring to the sentence about weak-head reduction, it can
  be deleted because it's discussed in Section 9 already}
Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use their semantic universe hierarchy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient for our purposes
(consistency and normalization).

>>>>

\subsection{Axioms}
Our Coq development assumes two axioms: functional extensionality
and propositional extensionality. The former is also required by
the Autosubst 2 libraries. Both axioms are known to be consistent
with Coq's metatheory.
%
These axioms bridge the gap between our mechanization and our informal
proofs. For example, in set theory, to show that two sets $[[S0]]$ and
$[[S1]]$ are equal, it suffices to show the extensional property that
$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
%
However, in Coq, sets of terms ($[[PowerSet STm]]$) are encoded as the 
type \texttt{tm
  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
not come with the extensionality property. Given two predicates $P$ and $Q$,
we cannot conclude that $P = Q$ when given a proof of
$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
predicate extensionality, an immediate corollary from functional
extensionality and propositional extensionality.

\subsection{Encoding the logical relation in Coq}
We next discuss specific details of the Coq encoding of the logical relation
presented in Section~\ref{sec:logreldep}.

@@

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, supported by the tools
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

@@

More recent work such as \citet{Abel12}
and \citet{abel2008betaeta} make their definitions more explicit and
precise and thus more directly encodable in proof assistants.
% Our logical relation resembles their definition of a
% semantic universe hierarchy.
Unlike \lang, their systems support type-directed $\eta$ laws. These
type-directed equational rules complicate the proof significantly
since the confluence of full reduction is not available until the
fundamental theorem is proven. Therefore,
instead of a logical predicate, a relational model with
syntactic type annotations is needed to remove the early reliance on
confluence, despite the fact that the underlying type theory is
intensional.

\citet{coquand:canonicity,coquand:prop} use a \emph{proof-relevant} logical
relation to prove canonicity and normalization for intensional
dependent type theories. % Intuitively, proof-relevant logical relations
% can be viewed as the set theoretic augmented with information
% about syntactic terms so the fundamental theorem can give us more
Compared
to the proof-irrelevant logical predicate we present, Coquand's proof
has the benefit that it does not rely on the confluence of reduction,
and can be easily extended to more $\eta$ laws without relying on a
complex relational model. Instead of working with concrete syntax,
Coquand works directly with an algebraic representation of dependent
type theory called category with families (CwF)~\citep{hofmann:cwf},
though the idea of proof-relevant logical relations for dependent
types can be expressed independently of category
theory~\citep{barras2012semantical}.

More abstractly, the proof technique by Coquand can be viewed as an
instance of a categorical construction called Artin Gluing
~\citep{kaposi:gluing}. Different metatheoretic
properties, such as normalization, parametricity, and
$\Pi$-injectivity can be obtained by gluing the initial CwF with the
corresponding CwF model. \citet{sterling:tait} introduces the notion of
\emph{Synthetic Tait Computability}, which allows further
simplification of such proofs by working in the internal language
formed by gluing.
However, proof-relevant logical relations are more demanding on the
expressive power of the metatheory. For example, the mechanization of
NbE by \citet{altenkirch:normalisation} relies on Quotient Inductive
Inductive Types (QIIT)~\citep{kaposi:qiit}.

% To support type-directed $\eta$ laws, \citet{Abel12,abel2008betaeta}
% use a partial equivalence relation to model their type system. The 
% prove the injectivity of $\Pi$ types cannot be derived 
% Furthermore, \citet{Abel12} and
% \citet{abel2008betaeta} use their semantic universe hierarchy as a
% measure to define Kripke-style logical relations, from which they
% derive the correctness of their conversion algorithms. In our work, we
% use the semantic universe hierarchy directly in our definition of
% semantic typing because it is sufficient for our purposes
% (consistency and normalization).
